---
title: "Complex Data Type Exercise: Immune Protein Sequences"
subtitle: "A beginner-friendly Biostrings analysis of FASTA protein data"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
---

## Goal

In this exercise, we will work with protein sequences stored in FASTA format.
FASTA files are common in biology because they can store one or many DNA, RNA,
or protein sequences in a simple text format.

We will use the `Biostrings` package to:

- read immune-related protein sequences from a FASTA file
- calculate the length of each protein sequence
- count amino acids in each sequence
- group amino acids by simple chemical property categories
- make exploratory plots

This is a practice analysis. The FASTA file below contains a small set of
immune-related protein sequence fragments so the code can run without
downloading external data.

## Load Packages

```{r}
# Biostrings is a Bioconductor package for reading and working with biological
# sequences. We use it here because protein FASTA files are complex data, not
# ordinary rectangular spreadsheets.

# If Biostrings is not installed yet, run these lines once in the Console:
# install.packages("BiocManager")
# BiocManager::install("Biostrings")

library(Biostrings)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Create a Small Practice FASTA File

```{r}
# In a real project, this path would point to a FASTA file downloaded from a
# database such as UniProt or NCBI.
#
# For this exercise, we create a small FASTA file ourselves. This makes the
# exercise reproducible because everyone starts with the same input file.

fasta_file <- "immune_protein_fragments.fasta"

# The lines that start with ">" are FASTA headers. Each header gives a name for
# the sequence that follows it.
#
# The other lines contain amino acid letters. For example:
# A = alanine, C = cysteine, D = aspartic acid, and so on.

practice_fasta <- c(
  ">IL2_fragment cytokine involved in T cell signaling",
  "MYRMQLLSCIALSLALVTNSAPTSSSTKKTQLQLEHLLLDLQMILNGINNYKNPKLTRMLTFKFYMPKKATELKHLQCLEEELKPLEEVLNLAQSKNFHLRPRDLISNINVIVLELKGSETTFMCEYADETATIVEFLNRWITFCQSIISTLT",
  ">IFNG_fragment cytokine involved in macrophage activation",
  "MQDPYVKEAENLKKYFNAGHSDVADNGTLFLGILKNWKEESDRKIMQSQIVSFYFKLFKNFKDDQSIQKSVETIKEDMNVKFFNSNKKKRDDFEKLTNYSVTDLNVQRKAIHELIQVMAELSPAAKTGKRKRSQMLFRGRRASQ",
  ">TNF_fragment inflammatory cytokine",
  "MSTESMIRDVELAEEALPKKTGGPQGSRRCLFLSLFSFLIVAGATTLFCLLHFGVIGPQREEFPRDLSLISPLAQAVRSSSRTPSDKPVAHVVANPQAEGQLQWLNRRANALLANGVELRDNQLVVPSEGLYLIYSQVLFKGQGCPSTHVLLTHTISRIAVSYQTKVNLLSAIKSPCQRETPEGAEAKPWYEPIYLGGVFQLEKGDRLSAEINRPDYLDFAESGQVYFGIIAL",
  ">CD4_fragment T cell surface receptor",
  "MNRGVPFRHLLLVLQLALLPAATQGKKVVLGKKGDTVELTCTASQKKSIQFHWKNSNQIKILGNQGSFLTKGPSKLNDRADSRRSLWDQGNFPLIIRLKIEDSDTYICEVEDQKEEVQLLVFGLTANSDTHLLQGQSLTLTLESPPGSSPSVQCRSPRGKNIQGGKTLSVSQLELQDSGTWTCTVLQNQKKVEFKIDIVVLAFQKASSIVYKKEGEQVEFSFPLAFTVEKLTGSGELWWQAERASSSKSWITFDLKNKEVSVKRVTQDPKLQMGKKLPLHLTLPQALPQYAGSGNLTLALEAKTGKLHQEVNLVVMRATQLQKNLTCEVWGPTSPKLMLSLKLENKEAKVSKREKAVWVLNPEAGMWQCLLSDSGQVLLESNIKVLPTWSTPVQPMALIVLGGVAGLLLFIGLGIFFCVRCRHRRRQAERMSQIKRLLSEKKTCQCPHRFQKTCSPI",
  ">CXCL8_fragment chemokine that recruits neutrophils",
  "MTSKLAVALLAAFLISAALCEGAVLPRSAKELRCQCIKTYSKPFHPKFIKELRVIESGPHCANTEIIVKLSDGRELCLDPKENWVQRVVEKFLKRAENS"
)

# Only write the file if it does not already exist. This avoids overwriting a
# file if you later replace it with your own sequences.
if (!file.exists(fasta_file)) {
  writeLines(practice_fasta, fasta_file)
}
```

## Read the FASTA File with Biostrings

```{r}
# readAAStringSet() reads amino acid sequences from a FASTA file.
# The result is an AAStringSet object, which is a Biostrings object designed for
# protein sequences.

immune_proteins <- readAAStringSet(fasta_file)

# Print the object so we can see how many sequences were read and their names.
immune_proteins
```

## Calculate Protein Sequence Lengths

```{r}
# width() gives the number of amino acids in each protein sequence.

protein_lengths <- tibble(
  protein = names(immune_proteins),
  length_aa = width(immune_proteins)
)

protein_lengths
```

```{r}
# A simple bar plot helps us compare sequence lengths.

ggplot(protein_lengths, aes(x = reorder(protein, length_aa), y = length_aa)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Protein Sequence Lengths",
    x = "Protein sequence",
    y = "Length in amino acids"
  ) +
  theme_minimal()
```

## Count Amino Acids

```{r}
# The standard protein alphabet has 20 common amino acid letters.

amino_acids <- c(
  "A", "R", "N", "D", "C",
  "Q", "E", "G", "H", "I",
  "L", "K", "M", "F", "P",
  "S", "T", "W", "Y", "V"
)

# alphabetFrequency() counts the letters in each sequence.
# baseOnly = TRUE keeps the output focused on standard amino acid letters.

aa_counts_matrix <- alphabetFrequency(
  immune_proteins,
  baseOnly = TRUE
)

# Convert the matrix to a data frame because data frames are easier to plot.

aa_counts <- as.data.frame(aa_counts_matrix) |>
  select(all_of(amino_acids)) |>
  mutate(protein = names(immune_proteins)) |>
  relocate(protein)

aa_counts
```

```{r}
# Change from wide format to long format.
# Long format is usually easier for ggplot because one row represents one
# protein and one amino acid.

aa_counts_long <- aa_counts |>
  pivot_longer(
    cols = all_of(amino_acids),
    names_to = "amino_acid",
    values_to = "count"
  )

aa_counts_long
```

```{r}
# Plot amino acid counts for each protein.

ggplot(aa_counts_long, aes(x = amino_acid, y = count)) +
  geom_col(fill = "darkseagreen4") +
  facet_wrap(~ protein) +
  labs(
    title = "Amino Acid Counts by Protein",
    x = "Amino acid",
    y = "Count"
  ) +
  theme_minimal()
```

## Calculate Amino Acid Percentages

```{r}
# Raw counts are useful, but longer proteins naturally have larger counts.
# Percentages make sequences of different lengths easier to compare.

aa_percent_long <- aa_counts_long |>
  left_join(protein_lengths, by = "protein") |>
  mutate(percent = 100 * count / length_aa)

aa_percent_long
```

```{r}
# This plot shows the percent of each amino acid in each sequence.

ggplot(aa_percent_long, aes(x = amino_acid, y = percent)) +
  geom_col(fill = "mediumpurple4") +
  facet_wrap(~ protein) +
  labs(
    title = "Amino Acid Composition by Protein",
    x = "Amino acid",
    y = "Percent of sequence"
  ) +
  theme_minimal()
```

## Group Amino Acids by Simple Properties

```{r}
# Amino acids can be grouped in many different ways.
# Here we use a simple beginner-friendly grouping.
#
# This grouping is only a broad summary. It is useful for exploration, but it
# does not replace detailed biochemical analysis.

aa_properties <- tibble(
  amino_acid = amino_acids,
  property_group = case_when(
    amino_acid %in% c("D", "E") ~ "Acidic",
    amino_acid %in% c("R", "H", "K") ~ "Basic",
    amino_acid %in% c("A", "V", "I", "L", "M", "F", "W", "Y", "P") ~ "Hydrophobic",
    amino_acid %in% c("S", "T", "N", "Q", "C", "G") ~ "Polar or small",
    TRUE ~ "Other"
  )
)

aa_properties
```

```{r}
# Join the property labels to the amino acid count table.
# Then add up counts within each protein and property group.

property_counts <- aa_counts_long |>
  left_join(aa_properties, by = "amino_acid") |>
  group_by(protein, property_group) |>
  summarise(
    count = sum(count),
    .groups = "drop"
  ) |>
  left_join(protein_lengths, by = "protein") |>
  mutate(percent = 100 * count / length_aa)

property_counts
```

```{r}
# A stacked bar plot gives a quick comparison of broad residue properties.

ggplot(property_counts, aes(x = protein, y = percent, fill = property_group)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Residue Property Groups",
    x = "Protein sequence",
    y = "Percent of sequence",
    fill = "Property group"
  ) +
  theme_minimal()
```

## Simple Questions to Consider

1. Which protein sequence is longest in this small example?
2. Which amino acids are most common across the proteins?
3. Do any proteins have a higher percentage of basic residues?
4. Do any proteins have a higher percentage of hydrophobic residues?
5. How might the results change if we used full-length reference sequences
   instead of short practice sequences?

## What This Exercise Shows

This analysis treats protein sequences as a complex data type. Unlike a CSV
file, a FASTA file is not already arranged as rows and columns. `Biostrings`
helps us read the sequences correctly, and then we can convert selected results
such as lengths and amino acid counts into tidy tables for plotting.
