---
title: "CDC Data Processing Exercise"
subtitle: "Influenza Vaccination Coverage"
author: "Yang Zeng"
output:
  html_document:
    toc: true
    toc_float: true
---

## Research question

How does influenza vaccination coverage vary across age groups, geography, and flu seasons?

## Data processing

The processing script imports the CDC data, converts unreported values to missing values, parses confidence intervals, validates numeric ranges, and creates comparable national and state subsets.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r load-data, include=FALSE}
source(
  "YangZeng-portfolio/cdcdata_exercise/cdcdata_exercise.R",
  local = knitr::knit_global()
)
```

## Missing data

The table reports missing vaccination estimates within each season, geography type, and demographic dimension.

```{r missing-data}
missing_summary
```

## National estimates by age group

This analysis uses national seasonal-influenza estimates from May, the end-of-season observation. Keeping the vaccine, geography, month, and dimension consistent prevents unrelated observations from being averaged together.

```{r national-summary}
national_age_summary
```

```{r national-age-plot, fig.width=10, fig.height=12}
national_age_plot
```

## State estimates

State records are identified using two-character FIPS codes. The summary describes the median and interquartile range across reporting states.

```{r state-summary}
state_season_summary
```

```{r state-distribution-plot, fig.width=10, fig.height=6}
state_distribution_plot
```
